The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-05-20 ms.Rmd
more tidying up of chunks, and putting Greg's sentences on newlines

```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
# set up knitr options
knitr::opts_chunk$set(fig.path = 'figs/',
               message = FALSE,
               dev = c('png'),
               cache = TRUE)
```

@@

SDM boom & growth of methods literature [@Barbosa2015]

Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly - fuelled by the availability of diverse modelling software and data. Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX). Most of the software that enables SDM research, and that of other computational sub-disciplines, focuses on the analytical tasks of science (such as modelling and statistics) rather than tasks that are essential to the scientific enterprise, such as communication and producing reproducible outputs.

This has presented computational science with a very modern version of the reproducibility problem as computational research is largely inaccessible (). If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g. Fig Share, Git Hub). The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules. The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).

@@

Modules are simply R function definitions, with a set of required inputs and outputs.
For example, the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is defined as:

```{r bioclimmodule, eval = FALSE}
Bioclim <- function (extent = c(-180, 180, -90, 90),
                     resolution = 10,
                     layers = 1:19) {

    if(!(resolution %in% c(2.5, 5, 10))){
      stop('only 2.5, 5 and 10 degree resolutions 
            are supported currently')
    }

    if (length(layers) < 1 |
        length(layers) > 19 |
        !all(layers %in% 1:19)) {
      stop ('layers must be a vector of integers between 
             1 and 19 indicating layers to obtain')
    }

    stopifnot(length(extent) == 4)
    stopifnot(all(is.numeric(extent)))

    world <- getData('worldclim', var = 'bio', res = resolution)
    world <- world[[layers]]
    cropped <- crop(world, extent(extent))
    return (cropped)
  }
```

@@

```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', message = FALSE}
Feng_Papes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = Chain(PerformanceMeasures, PrintMap))

>>>>

```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
# set up knitr options
knitr::opts_chunk$set(fig.path = 'figs/',
               message = FALSE,
               warning = FALSE,
               dev = c('png'),
               cache = TRUE)
```

@@

Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly [@Barbosa2015] - fuelled by the availability of diverse modelling software and data.
Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
Most of the software that enables SDM research, and that of other computational sub-disciplines, focuses on the analytical tasks of science (such as modelling and statistics) rather than tasks that are essential to the scientific enterprise, such as communication and producing reproducible outputs.

This has presented computational science with a very modern version of the reproducibility problem as computational research is largely inaccessible ().
If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g. Figshare, GitHub).
The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).

@@

Modules are simply R function definitions, with a set of required inputs and outputs.
For example the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is (excluding some error-handling code for brevity) defined as:

```{r bioclimmodule, eval = FALSE}
Bioclim <- function (extent = c(-180, 180, -90, 90),
                     resolution = 10,
                     layers = 1:19) {
    world <- getData('worldclim', var = 'bio', res = resolution)
    world <- world[[layers]]
    cropped <- crop(world, extent(extent))
    return (cropped)
  }
```

@@

```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE}
Feng_Papes <- 
  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                              extent = c(-130, -20, -60, 60)),
            covariate = Bioclim(extent = c(-130, -20, -60, 60),
                                layers = c(1:4, 6, 9, 10, 12, 15)),
              process = Chain(Clean, 
                              MESSMask,  
                              Background(n = 10000, bias = 200), 
                              Crossvalidate(k = 5)),
                model = MaxEnt,
               output = Chain(PerformanceMeasures, PrintMap))
